\mythesistitleNOTSYNPOSIS

### 2012-01-11 src/020chap2.tex
chapter 2 in edit still. see diff for changes.

\begin{definition}{\emph{Graphs that relate to
      COP.\cite[Def.~2.5]{d08phd}}} %
  \label{def:graphwithcop} %
  Let $G$ be a graph and $H$ be a bipartite graph.

  \begin{enumerate}[\ref{def:graphwithcop}--i.]
  \item $G$ is \emph{convex-round} if its adjacency matrix has the
    \CROP.
  \item \label{def::concave-round} $G$ is \emph{concave-round} if its
    augmented adjacency matrix has \CROP. \tnote{cite BHY00}
    \tnote[minor]{add CROP to glossary}
  \item $G$ is an \emph{interval graph} if its vertices can be mapped
    to intervals on the real line \stt two vertices are adjacent \iff
    their corresponding intervals overlap \tnote{cite Ben59, Haj57}.
    $G$ is an interval graph \iff its maximal clique matrix has COP
    \cite{fg65}\endnote{This follows \cite{gh64} which states that the
      maximal cliques of interval graph $G$ can be linearly ordered
      \stt for all $v \in V(G)$, cliques containing $v$ are
      consecutive in the ordering \cite[Th. 8.1]{mcg04}.}

    \begin{enumerate}[a.]
    \item $G$ is a \emph{unit interval graph} if it is an interval
      graph \stt all intervals have the same length.\endnotemark[2]
    \item $G$ is a \emph{proper interval graph} if it is an interval
      graph \stt no interval properly contains another.\endnote{The
        set of unit interval graphs and the set of proper interval
        graphs are the same} \tnote{cite {rob69,gar07} in endnote. pg
        33 dom}
    \end{enumerate}
 \item $G$ is a \emph{circular-arc graph} if its vertices can be
    mapped to a set of arcs on a circle \stt two vertices are adjacent
    \iff their corresponding arcs overlap.\tnote[pressing]{how is
      CO/ROP related?}
  \item $H$ is \emph{convex bipartite on columns (rows)} if its half
    adjacency matrix has \COP on rows (columns).%
    \label{def::convexbi}
  \item $H$ is \emph{biconvex bipartite} or \emph{doubly
      convex}\cite{yc95} if its half adjacency matrix has \COP on both
    rows and columns.
  \item $H$ is \emph{circular convex} if its half adjacency matrix has
    \CROP.
  \end{enumerate}
\end{definition}

Interval graphs and circular-arc graphs have a long history in
research. The interest around them is due to their very desirable
property that several problems that are NP-hard\tnote[minor]{all NPC
  problems are NPH yes?} on general graphs are not so in these graph
classes, \ie they are polynomial time solvable on them. In a similar
fashion, a lot of problems that are hard on general matrices have
efficient solutions on matrics with \COP or \CROP \cite[more citations
pg.\,33]{d08phd}.

Table~\ref{tab:graphmatrices} summarises the way these graphs are
characterized by their matrices having \COP or \CROP.
Our focus in this chapter (and thesis) is mainly \COP and 
having seen how useful \COP is in identifying or characterizing many
types of graphs, we will now see results that study recognition of \COP
in matrices in the following section.

@@

   \begin{tabular}[h]{lll}
     \hline
     \hline
     % 
     1899 & \parbox[t]{\tabparboxw}{ First mention of \COP} & \cite{p59}  \\

@@

@@

% \change{While it is remarkable that the first polynomial time
%   algorithm for \COP testing came in }{The first polynomial time
%   algorithm for \COP testing came in }\cite{fg65} \change{,}{.}
% \change{it is safe to say that the most}{A few years later, a very}
% significant result in understanding \COP came \remove{a few years
%   later} from Tucker which gave a combinatorial (negative)
% characterization of matrices with COP \cite{at72}.
A polynomial time algorithm for \COP testing was seen for the first
time in \cite{fg65} which uses overlapping properties of columns with
\un s. A few years later, a deeply significant result based on very
different ideas in understanding \COP came from Tucker which gave a
combinatorial (negative) characterization of matrices with COP
\cite{at72}. This result influenced most of the \COP results that
followed in the literature \annote[pressing]{including linear time
  algorithms}{did it? which ones?} for \COP recognition.

@@

Theorem~\ref{th:tuckeratfree} proves that convex bipartite graphs
cannot have {\em asteroidal triples}\endnote{If $G = (V,E)$ is a
  graph, a set of three vertices from $V$ form an {\em asteroidal
    triple} if between any two of them there exists a path in $G$ that
  does not contain any vertex from the closed neighborhood of the
  third vertex.} contained in the corresponding vertex
partition\endnote{The partition corresponds to columns (rows) if its
  half adjacency matrix has \COP columns (rows).}.
Theorem~\ref{th:tuckerforbidden} shows what are the structures in a
bipartite graph that force one of its vertex partitions to have
asteriodal triples -- in other words, it identifies the subgraphs that
prevent the graph from being convex bipartite.

@@

%\clearpage
\subsection{$PQ$ tree -- a linear COT algorithm}

Booth and Lueker in their paper \cite{bl76} gave the first linear
algorithm for \cop testing by way of an interval graph recognition
algorithm. Their result has close relations to the characterization of
interval graphs by \cite{gh64}. A graph $G$ is an interval graph \iff
all its maximal cliques can be linearly ordered \stt for any vertex
$v$ in $G$, all the cliques that $v$ is incident on are consecutive in
this order. Clearly, this means that the maximal clique incidence
matrix\footnote{
  Definition~\ref{def:graphmatrices}~\ref{def::maxcliquematrix}} must
have \COP on rows. This algorithm constructs a data structure called
\PQtree. A \PQtree represents all the \COP orderings of the matrix it
is associated with. \cite{bl76}'s algorithm uses the fact that if a
matrix has \COP, a \PQtree for it can be constructed. It is
interesting to note that aside from interval graph recognition and
\COP testing, \PQtree is also useful in other applications like
finding planar embeddings of planar graphs \cite{lec67,mcc04} and
recognizing \CROP in a matrix.

\begin{definition}[\emph{\PQtree \cite{bl76, mcc04}}]\\
  A \PQtree of matrix $M$ with \COP, is a tree with the following properties.
  \begin{enumerate}[i.]
    \singlespacing
  \item Each leaf uniquely represents a row (column) of $M$. The leaf
    order of the tree gives a \COP order for column (row)\endnote{Note
      that \COP order for column requires permutation of rows and vice
      versa.} for $M$.
  \item Every non-leaf node in the tree are labeled $P$ or $Q$.
  \item \label{def::nodep} The children of $P$ nodes are
    unordered. They can be permuted in any fashion to obtain a new
    \COP order for $M$.
  \item \label{def::nodeq} The children of $Q$ nodes are linearly
    ordered. Their order can be reversed to get obtain a new \COP
    order for $M$.
  \end{enumerate}
  \label{def:pqtree}
\end{definition}

@@

After the invention of \PQtrees, there has been several variants of
the same with either simpler procedure of construction, like \PCtree
\cite{wlh01} or a generalization of it, like
generalized \PQtree \cite{mcc04}. \PQRtree is another generalization
of \PQtree by \cite{mm96,mpt98}. There algorithm was improved by
\cite{tm05} \tnote{improved in terms of
  what?}. % describes an improved algorithm to
% build PQR trees.

\temptext{Expand a little more on PC trees and PQR trees.?}

We will see more of generalized \PQtree in Section~\ref{sec:surveycertalgo}.

\temptext{
  \begin{enumerate}[ -I-]
   \item Parallel algorithms [AS95, BS03, CY91]
  \end{enumerate}
}

\tnote[pressing]{ The results in cite~{bl76} on COT are based on the
   result that interval graphs are AT-free chordal graphs.
 -  Tucker's?}%

\tnote[pressing]{cite~{fg65} uses these results to give the first polynomial
  time algorithm for COT. WHICH RESULTS? \ \ \ \ \ check. how do they use it?}%

%\clearpage      % Print pending floats
\section{Optimization problems in COP}
\label{sec:surveycopopt}
\tnote{Expand on ref:sec:optcop}

%{\color{blue} \footnotesize
\temptext{
So far we have been concerned about matrices that have the consecutive
ones property. However in real life applications, it is rare that data
sets represented by binary matrices have COP, primarily due to the
noisy nature of data available. At the same time, COP is not arbitrary
and is a desirable property in practical data representation
\cite{co98,jkckv04,k77}. In this context, there are several
interesting problems when a matrix does not have COP but is ``close''
to having COP or is allowed to be altered to have COP. These are the
optimization problems related to a matrix which does not have
COP. Some of the significant problems are surveyed in this section.

\tnote{ -- sect 4.1 in cite:d08phd has many results
  surveyed. hardness results, approx. results. results are usually for
  a class of matrices $(a,b)$ where number {\un}s in columns and rows
  are restriced to $a$ and $b$ . -- problem of flipping at most $k$
  entries of $M$ to make it attain COP. this is NP complete
  cite:b75-phd}\tnote{(1) scite:lb62 showed that interval graphs
  are AT-free.  describe AT (2) show the close relationship b/w COP
  and graphs sec 2.2, pg 31} \cite{at72} showed that a matrix that
does not have COP have certain substructures that prevent it from
having COP. Tucker classified these forbidden substructures into five
classes of submatrices. This result is presented in the context of
convex bipartite graphs which \cite{at72} proved to be
AT-free\tnote{ check this up. give details. - doms'}. By
definition, convex bipartite graph have half adjacency matrices that
have COP on either rows or columns (graph is biconvex if it has COP on
both)\cite{d08phd}. A half adjacency matrix is a binary matrix
representing a bipartite graph as follows. The set of rows and the set
of columns form the two partitions of the graph. Each row node is
adjacent to those nodes that represent the columns that have {\un}s in
the corresponding row. \cite{at72} proves that this bipartite graph
has no asteroidal triple if and only if the matrix has COP and goes on
to identify the forbidden substructures for these bipartite
graphs. The matrices corresponding to these substructures are the
forbidden submatrices.

Once a matrix has been detected to not have COP (using any of the COT
algorithms mentioned earlier), it is naturally of interest to find out
the smallest forbidden substructure (in terms of number of rows and/or
columns and/or number of entries that are {\un}s). \cite{d08phd}
discusses a couple of algorithms which are efficient if the number of
{\un}s in a row is small. This is of significance in the case of
sparse matrices where this number is much lesser than the number of
columns. $(*,\Delta)${\em -matrices} are matrices with no restriction
on number of {\un}s in any column but has at most $\Delta$ {\un}s in
any row. {\sc Min COS-R (Min COS-C), Max COS-R (Max COS-C)} are
similar problems which deals with inducing COP on a matrix. In {\sc
  Min COS-R (Min COS-C)} the question is to find the minimum number of
rows (columns) that must be deleted to result in a matrix with COP.
In the dual problem {\sc Max COS-R (Max COS-C)} the search is for the
maximum number of rows (columns) that induces a submatrix with
COP. Given a matrix $M$ with no COP, \cite{b75-phd} shows that finding
a submatrix $M'$ with all columns\tnote{check if b75 deals with
  COP col or COP row. also is it any submatrix with k less than r rows
  or submatrix must have all columns?} but a maximum cardinality
subset of rows such that $M'$ has COP is NP complete. \cite{hg02}
corrects an error of the abridged proof of this reduction as given in
\cite{gj79}.  \cite{d08phd} discusses all these problems in detail
giving an extensive survey of the previously existing results which
are almost exhaustively all approximation results and hardness
results. Taking this further, \cite{d08phd} presents new results in
the area of parameterized algorithms for this
problem\tnote{elaborate - what are the results?}.

Another problem is to find the minimum number of entries in the matrix
that can be toggled to result in a matrix with COP.  \cite{v85}
discusses approximation of {\sc COP Augmentation} which is the problem
of changing of the minimum number of zero entries to {\un}s so that
the resulting matrix has COP. As mentioned earlier, this problem is
known to be NP complete due to \cite{b75-phd}. \cite{v85} also proves,
using a reduction to the longest path problem, \tnote{or is it a
  survey of another result?  check.} that finding a Tucker's forbidden
submatrix of at least $k$ rows is NP complete. \tnote{how is this
  different from booth's 75 result??}  \tnote{where should this go?
  cite|tz04 (approx submatrix with COP sparse matrices)}

\cite{jkckv04} discusses the use of matrices with almost-COP (instead
of one block of consecutive {\un}s, they have $x$ blocks, or {\em
  runs}, of consecutive {\un}s and $x$ is not too large) in the
storage of very large databases.  The problem is that of reordering of
a binary matrix such that the resulting matrix has at most $k$ runs of
{\un}s. This is proved to be NP hard using a reduction from the
Hamiltonian path problem.\tnote{Theorem 2.1 in jkckv} \tnote{(1) A
  connection of COP problem to the travelling salesman problem is also
  introduced. what does this mean? -- COP can be used as a tool to
  reorder $0.5T \le runs(M) \le T$. (2) The optimization version of
  the $k$-run problem, i.e. minimization of number of blocks of ones
  is proven to be NP complete by cite:k77}\tnote{are these two the
  same?} \tnote{what is the reduction?} \tnote{other problems similar
  to COP -- cite:ckl96 (ILP, circ ones, one drop) -- cite:th98
  (generalization of COP - minimax, biotonic column) Tucker}

%\section{Certifying Algorithms}
\section{COP for set systems}
\label{sec:surveycertalgo}
%\tnote{Expand on sec:appcertalgo} \tnote{ (certification McC04
%theme)}

Section~\ref{xxx} mentions how a binary matrix naturally maps to a
system of sets.  A set can be constructed for each column of matrix
with its elements being those row indices at which the column has \un
s. Thus the collection of sets corresponding each column of the matrix
forms a set system with universe as the set of all row indices of the
matrix.  This simple construction is formally described in
Definition~\ref{def:matrixsetsystem}.

@@

Thus the idea of \cop naturally maps to set sytems as well\endnote{As
  seen in Section~\ref{theCourseScheduleExample}} as
Definition~\ref{def:copforsets} states. This is equivalent to \COP for
matrices seen in Definition~\ref{def:copmatrix}.

@@

\temptext{ Certifying algorithm (Generalized PQ trees) [McC04] -- Move
  this to section on certifying algo.

@@

\section{COP in Relational Database Model}
\label{sec:surveyrdbm}
\tnote{Expand on sec:apprdbm} \tnote{(set systems theme)}

>>>>

\begin{definition}{\emph{Graphs that relate to
      COP.\cite[Def.~2.5]{d08phd}}} %
  \label{def:graphwithcop} %
  Let $G$ be a graph and $H$ be a bipartite graph.
  \begin{enumerate}[\ref{def:graphwithcop}--i.]
  \item $G$ is \emph{convex-round} if its adjacency matrix has the
    \CROP.
  \item \label{def::concave-round} $G$ is \emph{concave-round} if its
    augmented adjacency matrix has \CROP. \tnote{cite BHY00}
    \tnote[minor]{add CROP to glossary}
  \item $G$ is an \emph{interval graph} if its vertices can be mapped
    to intervals on the real line \stt two vertices are adjacent \iff
    their corresponding intervals overlap \tnote{cite Ben59, Haj57}.
    $G$ is an interval graph \iff its maximal clique matrix has COP
    \cite{fg65}\endnote{This follows \cite{gh64} which states that the
      maximal cliques of interval graph $G$ can be linearly ordered
      \stt for all $v \in V(G)$, cliques containing $v$ are
      consecutive in the ordering \cite[Th. 8.1]{mcg04}.}
    \begin{enumerate}[a.]
    \item $G$ is a \emph{unit interval graph} if it is an interval
      graph \stt all intervals have the same length.\endnotemark[2]
    \item $G$ is a \emph{proper interval graph} if it is an interval
      graph \stt no interval properly contains another.\endnote{The
        set of unit interval graphs and the set of proper interval
        graphs are the same} \tnote{cite {rob69,gar07} in endnote. pg
        33 dom}
    \end{enumerate}
 \item $G$ is a \emph{circular-arc graph} if its vertices can be
    mapped to a set of arcs on a circle \stt two vertices are adjacent
    \iff their corresponding arcs overlap.\tnote[pressing]{how is
      CO/ROP related?}
  \item $H$ is \emph{convex bipartite on columns (rows)} if its half
    adjacency matrix has \COP on rows (columns).%
    \label{def::convexbi}
  \item $H$ is \emph{biconvex bipartite} or \emph{doubly
      convex}\cite{yc95} if its half adjacency matrix has \COP on both
    rows and columns.
  \item $H$ is \emph{circular convex} if its half adjacency matrix has
    \CROP.
  \end{enumerate}
\end{definition}

Interval graphs\endnote { \cite{mcc04} cites that the problem of
  recognizing interval graphs has significance in molecular
  biology. Interestingly, in the late 1950s, before the structure of
  DNA was well-understood, Seymour Benzer was able to show that the
  intersection graph of a large number of fragments of genetic
  material was an interval graph \cite{sb59}. This was regarded as
  compelling evidence that genetic information was somehow arranged
  inside a structure that had a linear topology which we now know to
  be true\tnote{linear arrangement of DNA/genes?}.} and circular-arc
graphs have a long history in research.  The interest around them is
due to their very desirable property that several problems that are
NP-hard \tnote[minor]{all NPC problems are NPH yes?} on general
graphs, like finding a maximum clique or minimum coloring or
independent set, are polynomial time solvable in these graph classes
\cite{clrs01}.
In a similar fashion, a lot of problems that are hard on general
matrices have efficient solutions on matrices with \COP or \CROP
\cite[more citations pg.\,33]{d08phd}.

Table~\ref{tab:graphmatrices} summarises the way these graphs are
characterized by their matrices having \COP or \CROP.
Our focus in this chapter (and thesis) is mainly \COP and 
having seen how useful \COP is in identifying or characterizing many
types of graphs\endnote{\COP has several other applications
  \tnote{fill up.}}, we will now see results that study recognition of \COP
in matrices in the following section.

@@

   \begin{tabular}[h]{lll}
     \hline
     \hline
     % 
     1899 & \parbox[t]{\tabparboxw}{ First mention of \COP
       (archaeology)} & \parbox[t]{2cm}{\temptext{by petrie -- cite
         kendall 69 pacific journal of mathematics 1969 }}  \\

@@

The first polynomial time algorithm for \COP testing was by
\cite{fg65} which uses overlapping properties of columns with \un
s. Their result has close relations to the characterization of
interval graphs by \cite{gh64}. A graph $G$ is an interval graph \iff
all its maximal cliques can be linearly ordered \stt for any vertex
$v$ in $G$, all the cliques that $v$ is incident on are consecutive in
this order. Clearly, this means that the maximal clique incidence
matrix\endnote{
  Definition~\ref{def:graphmatrices}~\ref{def::maxcliquematrix}} must
have \COP on rows.

A few years later, a deeply significant result based on very
different ideas in understanding \COP came from Tucker which gave a
combinatorial (negative) characterization of matrices with COP
\cite{at72}. This result influenced most of the \COP results that
followed in the literature \annote[pressing]{including linear time
  algorithms}{did it? which ones?} for \COP recognition.

@@

% \change{While it is remarkable that the first polynomial time
%   algorithm for \COP testing came in }{The first polynomial time
%   algorithm for \COP testing came in }\cite{fg65} \change{,}{.}
% \change{it is safe to say that the most}{A few years later, a very}
% significant result in understanding \COP came \remove{a few years
%   later} from Tucker which gave a combinatorial (negative)
% characterization of matrices with COP \cite{at72}.

@@

Theorem~\ref{th:tuckeratfree} states that convex bipartite graphs
cannot have {\em asteroidal triples}\endnote{If $G = (V,E)$ is a
  graph, a set of three vertices from $V$ form an {\em asteroidal
    triple} if between any two of them there exists a path in $G$ that
  does not contain any vertex from the closed neighborhood of the
  third vertex.} contained in the corresponding vertex
partition\endnote{The partition corresponds to columns (rows) if its
  half adjacency matrix has \COP columns (rows).}.
Theorem~\ref{th:tuckerforbidden} lists the structures in a
bipartite graph that force one of its vertex partitions to have
asteriodal triples -- in other words, it identifies the subgraphs that
prevent the graph from being convex bipartite.

@@

%\clearpage
\subsection{Booth and Lueker's $PQ$ tree -- a linear COT algorithm}

Booth and Lueker in their paper \cite{bl76} gave the first linear
algorithm\endnote{Time complexity is $O\left(m+n+f\right)$ where $m
  \times n$ is the order of the input matrix and $f$ is the number of $\un$s in it.} for \cop testing while given a linear time interval graph
recognition algorithm by a simplification of \tnote{cite - A. Lempel,
  S. Even and I. Cederbaum, An Algorithm for Planarity Testing of
  Graphs, Theory of Graphs, ed., P. Rosenstiehl, Gordon and Breach,
  New York, (1967), 215-232.  }'s planarity test algorithm.
% Hsu in wlh01 said BL76 gave a planarity test - "the induced PQ-tree
% algorithm can considerably simplify Booth and Lueker’s modification
% of Lempel, Even and Cederbaum’s planarity test. " 
%
% McConnell in hm03 said the first linear algo for interval graph
% recog was by BL76 - "Benzer’s work and other applications of
% interval graphs motivated a search for eﬃcient algorithms for
% recognizing interval graphs, and for constructing a set of intervals
% to represent the graph when it is one [8]. A linear-time algorithm
% was given in 1976 by Booth and Lueker [2]." 
\cite{bl76} introduces a data structure called \PQtree and their \COP
testing algorithm is a constructive one that outputs a \PQtree if the
input has \COP. A \PQtree represents all the \COP orderings of the
matrix it is associated with. \cite{bl76}'s algorithm uses the fact
that if a matrix has \COP, a \PQtree for it can be constructed. It is
interesting to note that aside from interval graph recognition and
\COP testing, \PQtree is also useful in other applications like
finding planar embeddings of planar graphs \cite{lec67,mcc04} and
recognizing \CROP in a matrix.

\begin{definition}[\emph{\PQtree \cite{bl76, mcc04}}]\\
  A \PQtree of matrix $M$ with \COP on columns (rows), is a tree with the following properties.
  \begin{enumerate}[i.]
    \singlespacing
  \item Each leaf uniquely represents a row (column) of $M$. The leaf
    order of the tree gives a \COP order for column (row)\endnote{Note
      that \COP order for column requires permutation of rows and vice
      versa.} for $M$.
  \item Every non-leaf node in the tree is labeled $P$ or $Q$.
  \item \label{def::nodep} The children of $P$ nodes are
    unordered. They can be permuted in any fashion to obtain a new
    \COP order for $M$.
  \item \label{def::nodeq} The children of $Q$ nodes are linearly
    ordered. Their order can be reversed to obtain a new \COP
    order for $M$.
  \end{enumerate}
  \label{def:pqtree}
\end{definition}

@@

This algorithm is notoriously difficult to program. In the procedure
to change the \PQtree at each iteration, nodes are considered from
leaves to tree. At each node considered, it uses one of nine templates
to determine how the tree must be altered in the vicinity of this
node. Recognition of this template poses a difficult challenge in
terms of implementing it. Each template is actually a representative
of a larger class of similar templates, which must be dealt with
explicitly by a program\cite{mcc04}.

After the invention of \PQtrees, presumably due to the implementation
challenge it posed, there has been several variants of the same in the
literature, like \PCtree
\cite{sh99,wlh01,hm03}, generalized \PQtree \cite{km89,mcc04}, \PQRtree
\cite{mm96,mpt98} etc. Most of these are generalizations of \PQtree --
for instance, \PCtree is generalized to matrices with \CROP, \PQRtree and
generalized \PQtree are generalized to matrices and set systems with
or without \COP. \cite{km89} invented a modified form of \PQtree a simpler
incremental update of the tree only for recognizing interval graphs.
\cite{kr88} constructed efficient parallel algorithms for manipulating
PQ trees. \temptext{dom chapter 2 pg 40 Variations of PQTrees first
  para - summarize.}

In the next few following sections we will see some of these
variations.
%Their algorithm was improved by \cite{tm05}
%\tnote{improved in terms of what?}. 
% describes an improved algorithm to
% build PQR trees.

\subsection{\PCtree -- a generalization of \PQtree}

\PCtree is a generalization of \PQtree. It is a data structure that is
analogous to \PQtree but for matrices with \crop. \PCtree was
introduced by \cite{sh99} for the purpose of planarity testing where
this data structure represents partial embeddings of planar
graphs. % noted from wlh01 abstract.
In \cite{wlh01}, Hsu reintroduces \PCtree as a generalization of
\PQtree and shows how it simplifies \cite{bl76}'s planarity test by
making the \PQtree construction much less complicated. Later
\cite{hm03}, discovers that \PCtree is a representation of all \crop
orders of a matrix when it is unrooted. \PCtree presented in
\cite{wlh01} is rooted; however the construction of \PCtree is the
same in both results. The property of being unrooted is necessary in
order to use \PCtree as a data structure for encoding circular
ordering. Definition~\ref{def:pctree} defines \PCtree.

\begin{definition}
  \label{def:pctree}
  [\emph{\PCtree} \cite{wlh01,d08phd}.]  A \PCtree of matrix $M$ with
  \CROP on columns (rows), is a tree with the following properties.
  \begin{enumerate}[i.]
    \singlespacing
  \item Is unrooted -- thus it has (a) no parent child relationship
    between nodes (b) there is no left to right (or vice versa)
    ordering.
  \item Each leaf uniquely represents a row (column) of $M$. The leaf
    order of the tree gives a \CROP order for column (row) for
    $M$. Moreover, any sequence obtained by considering the leaves in
    clockwise or counter-clockwise order describes a \CROP order for
    $M$.
  \item Every non-leaf node in the tree is labeled $P$ or $C$.
  \item The neighbors of $P$ nodes can be permuted in any fashion to
    obtain a new \CROP order for $M$.
  \item \label{def::pcmirror} The tree can be changed by applying the following
    ``mirroring'' operation to obtain a new \CROP order for $M$. Root
    the \PCtree at a neighbor of a $C$-node, $v$ and mirror the
    subtree whose root is $v$ and finally unrooting the
    tree. Mirroring a subtree is done by putting the children of every
    node of the subtree in reverse order.
  \end{enumerate}
\end{definition}

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.4]{../img/pctree_hm03.png}

  \centering
  \includegraphics[scale=0.4]{../img/pctree_dom.png}
  \caption{\figtabsize \PCtree \temptext{ PLACEHOLDER 
IMGS \cite{hm03, d08phd}}}
  \label{fig:pctree}
\end{figure}

As a data structure when \PCtree is compared with \PQtree, the
differences are, (i) it is unrooted, (ii) it represents all \CROP order of a
matrix (iii) it has $C$ nodes instead of $Q$ nodes which can be
``mirrored'' (operation defined in Definition~\ref{def:pctree}~\ref{def::pcmirror}).
The algorithms of construction of \PQtree in \cite{bl76} and that of
\PCtree in \cite{wlh01, hm03} starkly differ since the latter is a
much simplified procedure.

% WLH01 abstract:
% [..] The original implementation of the PQ-tree algorithms by Booth and
% Lueker using nine templates in each bottom-up iteration is rather
% complicated. Also the complexity analysis is rather intricate. We give
% a very simple linear time PC-tree algorithm with the following
% advantages: (1) it does not use any template; (2) at each iteration,
% it does all necessary tree-modification operations in one batch and
% does not involve the node-by-node bottom-up matching; (3) it can be
% used naturally to test the circular ones property in matrices; (4) the
% induced PQ-tree algorithm can considerably simplify Booth and Lueker’s
% modification of Lempel, Even and Cederbaum’s planarity test.

\tnote[FYI]{ The results in cite~{bl76} on COT are based on the
   result that interval graphs are AT-free chordal graphs cite~{lb62}?
 No they don't seem related. this is the stupid bl lb confusion.}%

%\clearpage      % Print pending floats
%\section{Certifying Algorithms}
\subsection{Generalized \PQtree\ --  COP for set systems}
\label{sec:surveycertalgo}

Section~\ref{sec:motive} mentions how a binary matrix naturally maps
to a system of sets.  A set can be constructed for each column of
matrix with its elements being those row indices at which the column
has \un s. Thus the collection of sets corresponding each column of
the matrix forms a set system with universe as the set of all row
indices of the matrix.  This simple construction is formally described
in Definition~\ref{def:matrixsetsystem}.

@@

Thus the idea of \cop naturally maps to set sytems as well\endnote{As
  seen in Section~\ref{sec:courseschedule}} as
Definition~\ref{def:copforsets} states. This is equivalent to \COP for
matrices seen in Definition~\ref{def:copmatrix}.

@@

\tnote{What does n89 say? download paper from cornell website.}
Generalized \PQtree is a data structure defined in \cite{mcc04} for
arbitrary set systems. If the set system has \COP, the generalized
\PQtree is identical to its \PQtree.

An important observation made by \cite{mm96} is that if $\cF$ is a set
system with \COP then, aside from every set in $\cF$ being consecutive
(after applying the \COP order) the following sets must also be
consecutive for any $A, B \in \cF$ 
\begin{enumerate}
  \item The intersection $A \cap B$ 
  \item The union $A \cup B$ if $A \cap B \ne \emptyset$
  \item The relative complements $A \setminus B$ and $B \setminus A$
    if $B \nsubseteq A$ and  $A \nsubseteq B$ respectively.
  \item Additionally the sets $U$, singleton sets and the empty set $\emptyset$
    (collectively called {\em trivial sets} in \cite{mm96}) are consecutive in all orderings
    of the universe. $\emptyset$ is considered consecutive by convention.
\end{enumerate}

For any set system $\cF$, the smallest super set system containing
$\cF$ and the trivial sets that is closed under the above operations
is called the {\em weak closure} of $\cF$, denoted $\cW(\cF)$
\cite[Def.~3.2]{mcc04} \cite[Def.~2]{mm96}.

What is interesting is that the set of \COP permutations of $\cF$ is
the same as that of $\cW(\cF)$. \cite[Th.~3]{mm96}

The generalized \PQtree of $\cF$ is defined as the {\em decomposition
  tree} of its weak closure $\cW(\cF)$. Decomposition tree is the
inclusion tree (inclusion p.o. Hasse diagram) of the strong elements
of $\cW(\cF)$ with labels {\em prime, linear} and {\em degenerate}.\cite{mcc04}

\temptext{ Certifying algorithm (Generalized PQ trees) [McC04] --

@@

% \section{COP in Relational Database Model}
% \label{sec:surveyrdbm}
% \tnote{Expand on sec:apprdbm} \tnote{(set systems theme)}
\section{Optimization problems in COP}
\label{sec:surveycopopt}
\tnote{Expand on ref:sec:optcop}

\temptext{ So far we have been concerned about matrices that have the
  consecutive ones property. However in real life applications, it is
  rare that data sets represented by binary matrices have COP,
  primarily due to the noisy nature of data available. At the same
  time, COP is not arbitrary and is a desirable property in practical
  data representation \cite{co98,jkckv04,k77}. In this context, there
  are several interesting problems when a matrix does not have COP but
  is ``close'' to having COP or is allowed to be altered to have
  COP. These are the optimization problems related to a matrix which
  does not have COP. Some of the significant problems are surveyed in
  this section.

\tnote{ -- sect 4.1 in cite:d08phd has many results surveyed. hardness
  results, approx. results. results are usually for a class of
  matrices $(a,b)$ where number {\un}s in columns and rows are
  restriced to $a$ and $b$ . -- problem of flipping at most $k$
  entries of $M$ to make it attain COP. this is NP complete
  cite:b75-phd}%
\tnote{(1) scite:lb62 showed that interval graphs are chordal and
  AT-free.} %
\temptext{couple of lines refering to tucker's submatrices. refer
  earlier section.}

Once a matrix has been detected to not have COP (using any of the COT
algorithms mentioned earlier), it is naturally of interest

\begin{enumerate}
\item  to find out
the smallest forbidden substructure (in terms of number of rows and/or
columns and/or number of entries that are {\un}s). \cite{d08phd}
discusses a couple of algorithms which are efficient if the number of
{\un}s in a row is small. This is of significance in the case of
sparse matrices where this number is much lesser than the number of
columns.

\item $(*,\Delta)${\em -matrices} are matrices with no restriction
on number of {\un}s in any column but has at most $\Delta$ {\un}s in
any row. 
{\sc Min COS-R (Min COS-C), Max COS-R (Max COS-C)} are
similar problems which deals with {\em inducing COP} on a matrix.

\begin{enumerate}
\item In the dual problem {\sc Max COS-R (Max COS-C)} the search is for the
maximum number of rows (columns) that induces a submatrix with
COP.
\item In {\sc
  Min COS-R (Min COS-C)} the question is to find the minimum number of
rows (columns) that must be deleted to result in a matrix with COP.
\end{enumerate}

Given a matrix $M$ with no COP, \cite{b75-phd} shows that finding
a submatrix $M'$ with all columns\tnote{check if b75 deals with
  COP col or COP row. also is it any submatrix with k less than r rows
  or submatrix must have all columns?} but a maximum cardinality
subset of rows such that $M'$ has COP is NP complete. \cite{hg02}
corrects an error of the abridged proof of this reduction as given in
\cite{gj79}.  \cite{d08phd} discusses all these problems in detail
giving an extensive survey of the previously existing results which
are almost exhaustively all approximation results and hardness
results. Taking this further, \cite{d08phd} presents new results in
the area of parameterized algorithms for this
problem\tnote{elaborate - what are the results?}.

\item Another problem is to find the minimum number of entries in the matrix
that can be toggled to result in a matrix with COP.  \cite{v85}
discusses approximation of {\sc COP Augmentation} which is the problem
of changing of the minimum number of zero entries to {\un}s so that
the resulting matrix has COP. As mentioned earlier, this problem is
known to be NP complete due to \cite{b75-phd}. \cite{v85} also proves,
using a reduction to the longest path problem, \tnote{or is it a
  survey of another result?  check.} that finding a Tucker's forbidden
submatrix of at least $k$ rows is NP complete. \tnote{how is this
  different from booth's 75 result??}  \tnote{where should this go?
  cite|tz04 (approx submatrix with COP sparse matrices)}

\item \cite{jkckv04} discusses the use of matrices with almost-COP (instead
of one block of consecutive {\un}s, they have $x$ blocks, or {\em
  runs}, of consecutive {\un}s and $x$ is not too large) in the
storage of very large databases.  The problem is that of reordering of
a binary matrix such that the resulting matrix has at most $k$ runs of
{\un}s. This is proved to be NP hard using a reduction from the
Hamiltonian path problem.\tnote{Theorem 2.1 in jkckv} \tnote{(1) A
  connection of COP problem to the travelling salesman problem is also
  introduced. what does this mean? -- COP can be used as a tool to
  reorder $0.5T \le runs(M) \le T$. (2) The optimization version of
  the $k$-run problem, i.e. minimization of number of blocks of ones
  is proven to be NP complete by cite:k77}\tnote{are these two the
  same?} \tnote{what is the reduction?} \tnote{other problems similar
  to COP -- cite:ckl96 (ILP, circ ones, one drop) -- cite:th98
  (generalization of COP - minimax, biotonic column) Tucker}

\end{enumerate}
